Testing and Isolation Efficacy:\\Insights from a Simple Epidemic Model

### 2021-06-22 SIR_manuscript.tex
revisions to Results

The basic reproduction number, $\Rnum$, was calculated by using the next-generation matrix method \citep{van2002reproduction}. We present $\Rnum$ as
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} \left(1-\Delta\right), 
\end{equation}
where the term $\beta/\gamma$ is the classical basic reproduction number for a simple SIR model \citep{keeling2011modeling}, and $0 < \Delta < 1$ is the proportional reduction \david{actually $1-\Delta$ is the proportional reduction} due to testing and isolation processes, calculated as 
\begin{equation}
  \label{eq:del4a}
  \Delta= \frac{1}{C N_0}\big(C_1 S\_u^*+(C_2(1-\theta\_w)+C \theta\_w)S\_n^*\big),
\end{equation}
where
\begin{align}
\label{eq:C12}
C_{\phantom 1} &= (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big),\\
C_1 &= (\omega+\gamma)(\theta\_w \gamma+\theta\_c \omega p_I) \hat F_I,\\
C_2 &= \Big( \omega\gamma(1+p_I)\hat F_I+\gamma^2(\omega+\gamma+\hat F_I)\Big)\theta\_w + \omega^2p_I\hat F_I\theta\_c.
\end{align}
(\appref{R0} gives a detailed derivation of these expressions.)
This explicit formula enables us to study the effects of testing and isolation parameters on $\Rnum$ both analytically and via numerical solutions.
We are specifically interested in parameters that could be manipulated by public health policy: isolation efficacy, $\theta\_c$ and $\theta\_w$; \percap testing intensity, $\rho$; and the rate of test return, $\omega$. In particular, we look at the partial derivatives of $\Delta$ with respect to these parameters (\appref{rho}). 
We derived general expressions for these derivatives. However, we analyzed the effect of $\omega$ on $\Delta$ for the special case of low testing intensity. Specifically, by making the restriction $\rho \ll 1$, we are able to Taylor-expand $\Delta$ at $\rho=0$, use the linear approximation with respect to $\rho$ and analyze the resulting simplified derivatives to illustrate a surprising non-monotonic relationship between $\Delta$ and $\omega$.

@@

We can use the formula for $\Rnum$ \eqref{R0} to make a number of straightforward conclusions about parameters whose effects on $\Rnum$ are monotonic, i.e. when the partial derivative of $\Delta$ has a consistent sign
(see Appendices).

@@

\item \label{p1:eta} Increasing isolation efficacy for waiting ($\theta\_w$) and confirmed-positive ($\theta\_c$) individuals always decreases \Rnum (Eqs.~\ref{eq:del4_theta},~\ref{eq:d_del4_thetac},~\ref{eq:d_del4_thetaw});
\item \label{p1:rho} Higher testing intensity $\rho$ decreases $\Rnum$ if
%% \begin{enumerate} \item
testing is random (all $w_X$ equal) or testing intensity ($\rho$) is small (Eq.~\ref{eq:dd3dr}).
%% \end{enumerate}
\item \label{p1:omega} Increasing the rate of test return ($\omega$) always decreases \Rnum if waiting individuals do not isolate ($\theta\_w=0$) (Eq.~\ref{eq:dlindo}).
\item \label{p1:w} Increasing testing focus, i.e. changing the testing weights from random ($w_S=w_I$) toward targeted  ($w_S<w_I$), always decreases \Rnum (Eq.~\ref{eq:d_del_wis}).
\end{enumerate}

\newpage
\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf Effectiveness of testing and isolation in reducing $\Rnum$ at low \percap testing intensity.}
Numerical evaluation of the effectiveness of control ($\Delta$: eq.~\ref{eq:del4}), over a range of testing and isolation parameters. Parameter values (Table~\ref{tab:params}):
$\beta=\betaparam$ 1/day, $1/\gamma= \invgammaparam$ days (baseline $\Rnum=\Rnumparam$, $r=\rparam$); $\omega \in [1/12,2]$ 1/day;  $\rho \in [0,0.013]$ 1/day per capita; $\theta\_w$ and $\theta\_c$ vary between 0 (no effect of isolation) and 1 (complete elimination of transmission); $p_S=0$, $p_I=1$ and $p_R=0.5$. Only parameter sets where $\theta\_c \geq \theta\_w$ (confirmed-positive individuals isolate more effectively than waiting individuals) are shown; the alternative case, $\theta\_w > \theta\_c$, is unrealistic. Contours of $\Delta$ are plotted for (a) random testing ($w_S=w_I=w_R=1$) and (b) targeted testing ($w_S=0.3$; $w_I=w_R=1$). 
}
\label{pan}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random2.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI2.pdf}
\caption{}
\end{subfigure}
\caption{
  {\bf Effectiveness of testing and isolation in reducing $\Rnum$ at high \percap testing intensity.}
  Numerical evaluation of the effectiveness of control ($\Delta$: eq.~\ref{eq:del4}), over a range of testing and isolation parameters. Parameters as in \fref{pan} except: $\omega \in [1/5,2]$ 1/day, $\rho \in [0,1/5)$ 1/day. As in \fref{pan}, subplots show (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$.
}
\label{pan2}
\end{figure}

@@

\begin{itemize}
\item We would generally expect increasing testing delays to increase \Rnum, thus decreasing effectiveness of control $\Delta$. This is in fact what happens when waiting individuals do not isolate ($\theta\_w =0$, top row of \fref{pan}) --- as we move to the right within each plot in this row, $\Delta$ decreases.
However, when waiting individuals isolate ($\theta\_w>0$), we more often see the opposite effect: longer testing delays lead to a greater control effect $\Delta$. This is because people waiting for negative tests are assumed to continue to isolate; this applies both to susceptibles and to people who became infected while waiting for negative test results. This effect outweighs the effect of confirmed individuals isolating except when this isolation parameter $\theta\_c$ is substantially bigger than $\theta\_w$. This result depends on the idea that, all else equal, people who have to wait longer for test results isolate at the same level (but for a longer time) as they would if the wait were shorter.
\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). This relationship can be reversed, however, at very high testing intensities. This can only occur when testing is targeted, and $\theta\_w$ is relatively small (\fref{pan2}(b), right three panels of top row). It is theoretically possible for increasing testing intensity to \emph{increase} \Rnum because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; if these people become infected while waiting, they will need to wait for their negative test result before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.

>>>>

The basic reproduction number, $\Rnum$, was calculated by using the next-generation matrix method \citep{van2002reproduction}. We present $\Rnum$ as
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} \left(1-\Delta\right), 
\end{equation}
where the term $\beta/\gamma$ is the classical basic reproduction number for a simple SIR model \citep{keeling2011modeling}, and $0 < \Delta < 1$ is the proportional reduction \david{actually $1-\Delta$ is the proportional reduction; you should refer to this as the effectiveness of control, as you do below} due to testing and isolation processes, calculated as 
\begin{equation}
  \label{eq:del4a}
  \Delta= \frac{1}{C N_0}\big(C_1 S\_u^*+(C_2(1-\theta\_w)+C \theta\_w)S\_n^*\big),
\end{equation}
where
\begin{align}
\label{eq:C12}
C_{\phantom 1} &= (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big),\\
C_1 &= (\omega+\gamma)(\theta\_w \gamma+\theta\_c \omega p_I) \hat F_I,\\
C_2 &= \Big( \omega\gamma(1+p_I)\hat F_I+\gamma^2(\omega+\gamma+\hat F_I)\Big)\theta\_w + \omega^2p_I\hat F_I\theta\_c.
\end{align}
(\appref{R0} gives a detailed derivation of these expressions.)
This explicit formula enables us to study the effects of testing and isolation parameters on $\Rnum$ both analytically and via numerical solutions.
We are specifically interested in parameters that could be manipulated by public health policy: isolation efficacy, $\theta\_c$ and $\theta\_w$; \percap testing intensity, $\rho$; and the rate of test return, $\omega$. In particular, we look at the partial derivatives of $\Delta$ with respect to these parameters (\appref{rho}). 
We derived general expressions for these derivatives. However, we analyzed the effect of $\omega$ on $\Delta$ for the special case of low testing intensity. Specifically, by making the restriction $\rho \ll 1$, we are able to Taylor-expand $\Delta$ at $\rho=0$, use the linear approximation with respect to $\rho$ and analyze the resulting simplified derivatives to illustrate a surprising non-monotonic relationship between $\Delta$ and $\omega$.

@@

We can use the formula for $\Rnum$ \eqref{R0} to make a number of straightforward inferences about parameters that affect $\Rnum$ monotonically, i.e., for which the associated partial derivative of $\Delta$ always has the same sign (see Appendices).

@@

\item \label{p1:eta} Increasing isolation efficacy for waiting ($\theta\_w$) and confirmed-positive ($\theta\_c$) individuals always decreases \Rnum (Eqs.~\ref{eq:del4_theta},~\ref{eq:d_del4_thetac},~\ref{eq:d_del4_thetaw});
\item \label{p1:rho} Higher testing intensity $\rho$ decreases $\Rnum$ if
%% \begin{enumerate} \item
testing is random (all $w_X$ equal) or testing intensity ($\rho$) is small (Eq.~\ref{eq:dd3dr}).
%% \end{enumerate}
\item \label{p1:omega} Increasing the rate of test return ($\omega$) always decreases \Rnum if waiting individuals do not isolate ($\theta\_w=0$) (Eq.~\ref{eq:dlindo}).
\item \label{p1:w} Increasing testing focus, i.e., changing the testing weights from random ($w_S=w_I$) toward targeted  ($w_S<w_I$), always decreases \Rnum (Eq.~\ref{eq:d_del_wis}).
\end{enumerate}

\newpage
\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf Effectiveness of testing and isolation in reducing $\Rnum$ at low \percap testing intensity ($\rho$).}
Numerical evaluation of the effectiveness of control ($\Delta$: eq.~\ref{eq:del4}\david{wrong eq ref}), over a range of testing and isolation parameters. Parameter values (Table~\ref{tab:params}):
$\beta=\betaparam/$day, $1/\gamma= \invgammaparam$ days (baseline $\Rnum=\Rnumparam$, $r=\rparam$); $\omega \in [1/12,2]/$day;  $\rho \in [0,0.013]/$day per capita; $\theta\_w$ and $\theta\_c$ vary between 0 (no effect of isolation) and 1 (complete elimination of transmission); $p_S=0$, $p_I=1$ and $p_R=0.5$. Only parameter sets where $\theta\_c \geq \theta\_w$ (confirmed-positive individuals isolate more effectively than waiting individuals) are shown; the alternative case, $\theta\_w > \theta\_c$, is unrealistic. Contours of $\Delta$ are plotted for (a) random testing ($w_S=w_I=w_R=1$) and (b) targeted testing ($w_S=0.3$; $w_I=w_R=1$). 
}
\label{pan}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random2.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI2.pdf}
\caption{}
\end{subfigure}
\caption{
  {\bf Effectiveness of testing and isolation in reducing $\Rnum$ at high \percap testing intensity.}
  Numerical evaluation of the effectiveness of control ($\Delta$: eq.~\ref{eq:del4}\david{again wrong eq ref}), over a range of testing and isolation parameters. Parameters as in \fref{pan} except: $\omega \in [1/5,2]/$day, $\rho \in [0,1/5)/$day. As in \fref{pan}, subplots show (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$.
}
\label{pan2}
\end{figure}

@@

\begin{itemize}
\item We would generally expect increasing testing delays to increase \Rnum, thus decreasing effectiveness of control $\Delta$. This is in fact what happens when waiting individuals do not isolate ($\theta\_w =0$, top row of \fref{pan}) --- as we move to the right within each plot in this row, $\Delta$ decreases.
However, when waiting individuals isolate ($\theta\_w>0$), we more often see the opposite effect: longer testing delays lead to a greater control effect $\Delta$ (reduced $\Rnum$). The reason is that people waiting for negative tests are assumed to continue to isolate; this applies both to susceptibles and to people who became infected while waiting for negative test results. This effect outweighs the effect of confirmed individuals isolating, except when this isolation parameter ($\theta\_c$) is substantially greater than $\theta\_w$. This result depends on the idea that, all else equal, people who have to wait longer for test results isolate at the same level (but for a longer time) as they would if the wait were shorter.
\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). However, this relationship can be reversed at very high testing intensities (provided testing is targeted, and $\theta\_w$ is relatively small; \fref{pan2}(b), right three panels of top row). It is theoretically possible for increasing testing intensity to \emph{increase} \Rnum because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; if these people become infected while waiting, they will need to wait for their negative test result before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.
